The Second Data Release of the INT Photometric H$\alpha$ Survey of the Northern Galactic Plane (IPHAS DR2)

### 2014-04-16 paper/iphas-dr2.tex

\begin{figure*}
        \includegraphics[width=1\linewidth]{figures/footprint/footprint_large.jpg}
        \caption{Survey area showing the footprints
        of all the quality-approved IPHAS fields
        which have been included in this data release.
        The area covered by each field has been coloured black
        with a semi-transparent opacity of 20 per cent,
        such that regions where fields overlap are darker.
        The IPHAS strategy is to observe each field twice
        with a small offset,
        and hence the vast majority of the area 
        is covered twice (dominant grey colour).
        There are small overlaps between all the neighbouring fields
        which can be seen as a honeycomb pattern
        of dark grey lines across the survey area.
        Regions with incomplete data are apparent as white gaps (no data) 
        or in light grey (indicating that one offset is missing).
        The dark vertical strip near $\ell \simeq$125\degr\ 
        is an arbitrary consequence of the tiling pattern,
        which was populated starting from 0h in Right Ascension.}
        \label{fig:footprint}
\end{figure*}

@@

(3) \emph{Seeing.} 
The original survey goal was to obtain data 
at a resolution better than 1.7~arcsec,
as evaluated by measuring the average PSF Full Width at Half Maximum (FWHM)
across the detectors.
% Stating the obvious ...a new insertion? which is a proxy for the image quality
% arising from the combined effects of atmospheric and dome seeing.
This target is currently attained across 86 per cent of the footprint.
To increase the sky area offered by the data release slightly,
we have decided to accept data obtained with FWHM up to 2.5~arcsec.
Fig.~\ref{fig:seeing} presents the distribution
of the PSF FWHM for the approved fields.
In the $r$-band, 90 per cent is better than 1.5~arcsec,
50 per cent is better than 1.1~arcsec,
and 10 per cent is better than 0.8~arcsec.
% slipped in earlier (the WFC pixel scale is 0.33~arcsec/px).
In \S\ref{sec:catalogue} we will explain
that the photometry compiled in the source catalogue
is normally derived from the field with the
best-available seeing for a given object,
and that the FWHM measurement
is available as a column in the catalogue.

@@

Unlike the broadbands, 
standard-star photometry is not available in the literature 
for the \ha\ passband
and hence there is no formally recognised flux scale 
for it.
We can specify here, however, 
that the detected flux for Vega in the
IPHAS \ha\ filter is 3.14 magnitudes
less than the flux captured by the much broader
$r$ band which includes the \ha\ band.
Hence to assure zero colour relative to the broadbands,
we set the default zeropoint for the narrowband to be:
\begin{equation}
\textsc{magzpt}_{H\alpha} = \textsc{magzpt}_r - 3.14.
\label{eqn:zpha}
\end{equation}
For reference, Table~\ref{tbl:flux} details the flux of Vega
in the IPHAS filter system.
Data on the throughput curves of the bands can be obtained from
the Isaac Newton Group website\footnote{http://catserver.ing.iac.es/filter/list.php?instrument=WFC where the
filters are named WFCH6568, WFCSloanR and WFCSloanI}.

\begin{table}
    \caption{Mean monochromatic flux of Vega in the IPHAS filter system,
             defined as 
             $\langle f_\lambda\rangle~=~\int{f_\lambda(\lambda) S(\lambda) \lambda d \lambda} / \int{S(\lambda)\lambda d \lambda}$,
             where $S(\lambda)$ is the photon response function 
             (which includes atmospheric transmission,
             filter transmission
             and CCD response)
             and $f_\lambda(\lambda)$ is the CALSPEC SED for Vega
             \citep{Bohlin2014}.
             For reference, we also provide the 
             filter equivalent width EW~=~$\int S(\lambda) d\lambda$,
             the mean photon wavelength 
             $\lambda_0~=~\int{S(\lambda)\lambda d\lambda} / \int{S(\lambda)d\lambda}$,
             and the pivot wavelength
             $\lambda_p~=~\sqrt{\int{S(\lambda)\lambda d\lambda} / \int{\frac{S(\lambda)}{\lambda} d\lambda}}$.
             These notations follow the definitions by \citet{Bessell2012}.
             After multiplying $\langle f_\lambda\rangle$ by the EW,
             we find that the detected flux for Vega in \ha\ is 3.14~magnitudes less
             than that received in $r$, 
          }
    \label{tbl:flux}
    \begin{center}
    \begin{tabular}{lccrc}
    \toprule
     Filter  & $\langle f_\lambda \rangle$  & \multicolumn{1}{c}{EW} &$\lambda_0$ & $\lambda_p$    \\
             & [erg\,cm$^{-2}$\,s$^{-1}$\,\AA$^{-1}$] & [\AA] & [\AA] & [\AA] \\
    \midrule
$r$ &  $2.47 \times 10^{-9}$  &  785.6 &  6223  &   6211   \\
\ha &  $1.81 \times 10^{-9}$ &   59.6 &  6568  &   6568    \\
$i$  &  $1.30 \times 10^{-9}$  &  759.9  &  7674  &   7661   \\
    \bottomrule
    \end{tabular}
    \end{center}
\end{table}

@@

Despite the best efforts made to obtain a nightly calibration,
large surveys naturally possess field-to-field variations
due to atmospheric changes during the night
and imperfections in the pipeline or the instrument
(e.g. the WFC is known to suffer from sporadic errors
in the timing of exposures).
This is demonstrated in Fig.~\ref{fig:ccd_before},
where we show the combined colour-colour diagram
for the set of 2\,801 quality-approved fields
which are located towards the Galactic anti-centre
($160\degr < \ell < 200\degr$).
Despite the fact that all the fields included in the diagram
have been quality-approved,
the main locus of stars is poorly defined
due to the presence of calibration errors,
which regularly exceed 0.1~mag.
For comparison, Fig.~\ref{fig:ccd_after}
shows the identical diagram
obtained after re-calibrating.

\begin{figure}
\captionsetup[subfigure]{labelformat=empty}
\begin{subfigure}[b]{\linewidth}
\centering
\includegraphics[width=\textwidth]{figures/caldiagram/ccd-uncalibrated.pdf}
\caption{}
\label{fig:ccd_before}
\end{subfigure}
\begin{subfigure}[b]{\linewidth}
\centering
\includegraphics[width=\textwidth]{figures/caldiagram/ccd-calibrated.pdf}
\caption{}
\label{fig:ccd_after}
\end{subfigure}
\caption{IPHAS $(r-\ha,\ r-i)$ colour-colour diagram for an area of 400~deg$^2$, plotted before (panel a) and after (panel b)
re-calibration.
Both figures were created by combining the stars detected
across the entire set of 2\,801 quality-approved fields which are located
towards the Galactic anti-centre  ($160\degr < \ell < 200\degr$).
The diagrams are plotted as 2D-histograms
which show the density of sources
in bins of 0.01-by-0.01~mag;
bins containing 1 to 200 sources are coloured red,
while bins containing more than 400 sources are bright yellow.
We have included all stars which are brighter than $r<18$
and are classified as `very reliable'
(these criteria will be explained in \S\ref{sec:qualitycriteria}).}
\end{figure}

The calibration variations need to be corrected for 
during a global re-calibration procedure.
Notable past examples include the global re-calibration 
of 2MASS \citep[][]{Nikolaev2000},
SDSS \citep[][]{Padmanabhan2008}
and the Panoramic Survey Telescope 
And Rapid Response System survey \citep[Pan-STARRS;][]{Schlafly2012},
which all achieved photometry 
that is globally consistent to within 0.01--0.02~mag.

Surveys which observe identical stars at different epochs
can use the repeat measurements to ensure a homogeneous calibration.
For example, 2MASS attained its global calibration
by observing six standard fields each hour, 
allowing zeropoint variations to be tracked 
over very short timescales \citep{Nikolaev2000}.
Alternatively, the SDSS and PanSTARRS surveys could benefit
from revisiting regions in their footprint to 
carry out a so-called \emph{ubercalibration}\footnote{`Ubercalibration'
refers to the name of the code used to re-calibrate SDSS photometry
using repeat observations. 
It is an anglicised version of the German word `\"uberkalibration',
which was reportedly chosen because the initial authors both
had German-sounding names \citep{Finkbeiner2010}.} procedure,
in which multiple measurements of stars at different epochs
are used to fit the calibration parameters
\citep[e.g.][]{Ivezic2007,Padmanabhan2008,Schlafly2012}.

@@

We have found the information contained
in these overlap regions to be insufficient
to constrain the calibration parameters.
This is because photometry at the extreme edges of the WFC
-- where neighbouring fields overlap -- 
is prone to systematics at the level of 1--2~per cent.
The cause of these errors is thought to include 
the use of twilight sky flats in the pipeline,
which are known to be imperfect for calibrating stellar photometry 
due to stray light and vignetting \citep[e.g.][]{Manfroid1995}.
Moreover, the illumination correction in the overlap regions
is more affected by a radial geometric distortion in the WFC,
which causes the pixel scale to increase as the edges 
are approached \citep{Gonzalez-Solares2011}.
Although these systematics are reasonably small within a single field,
they can accumulate during the calibration process, causing artificial zeropoint gradients 
across the survey unless controlled by other external constraints.

@@

We have been able to benefit from APASS
(http://www.aavso.org/apass)
to bring most of the survey 
onto a uniform scale.
Since 2009,
APASS has been using two 20~cm-astrographs
to survey the entire sky down to $\sim$17th magnitude
in five filters which include Sloan $r$ and $i$ \citep{Henden2012}.
The most recent catalogue available 
at the time of preparing this work was APASS DR7,
which provides a good coverage across $\sim$half
of the IPHAS footprint.
The overlap regions are shown in Fig.~\ref{fig:apass_r}.
The photometric accuracy of APASS is currently estimated 
to be at the level of 3 per cent,
%The private communication from Hended is no longer needed :-)
which is significantly better 
than the original nightly calibrations of IPHAS
which are accurate to $\sim$10 per cent (Tab.~\ref{tbl:offsets_before}).
APASS achieves its uniform accuracy 
by measuring each star at least two times in photometric conditions,
along with ample standard fields,
benefiting from the large 3-by-3 degrees field of view of its detectors.

@@

The resulting set of $\sim$220\,000 cross-matched stars were then used 
to derive APASS-to-IPHAS magnitude transformations
using a linear least-squares fitting routine, 
which iteratively removed $3\sigma$-outliers to improve the fit.
The solution converged to:
\begin{align} 
r_{\rm IPHAS} = r_{\rm APASS} - 0.121 + 0.032(r-i)_{\rm APASS} \label{eqn:apass_r} \\
i_{\rm IPHAS} = i_{\rm APASS} - 0.364 + 0.006(r-i)_{\rm APASS} \label{eqn:apass_i}
\end{align}
The root mean square (rms) residuals of these transformations 
are 0.041 and 0.051, respectively.
The small colour terms in the equations
indicate that the IPHAS and APASS broadband filters 
are very similar.
The transformations include a large fixed offset,
but this is simply due to the fact that 
APASS magnitudes are given in the AB system
and IPHAS uses Vega-based magnitudes.
%I wonder if the next bit is stating the too-obvious?
%Have shortened it a bit.
Separate transformations were derived for sightlines 
with varying extinction properties to investigate the robustness
of the transformations with respect to different reddening regimes.
This sensitivity was found to be insignificant, in keeping with
the scarcity of heavily-reddened objects at $r<16$.

Having transformed APASS magnitudes into the IPHAS system,
we then computed the median magnitude offset 
for each field which contained at least 30 cross-matched stars.
This was achieved for 48 per cent of our fields (shown in Figs.~\ref{fig:apass_r} and \ref{fig:apass_i}).
The offsets follow a near-Gaussian distribution
centred around $0.014\pm0.104$~mag in $r$ and $0.007\pm0.108$~mag in $i$
(Table~\ref{tbl:offsets_before}).
A total of 4596 fields showed a systematic offset
exceeding $\pm$0.03~mag in either $r$ or $i$.

@@

We then applied the most important step in our calibration scheme,
which is to adjust the provisional zeropoints of these 4596 aberrant fields
such that their offset is brought to zero.
This allowed the mean IPHAS-to-APASS offset 
to be brought down to $0.000\pm0.011$~mag in both $r$ and $i$
(Table~\ref{tbl:offsets_after}).
The procedure of fitting magnitude transformations and
correcting the IPHAS zeropoints was repeated a few times to ensure 
convergence, which was closely approached after the first iteration.

@@

At the time of writing, the APASS catalogue did not provide 
sufficient coverage for 7359 of our fields.
Fortunately, these fields are located mainly 
at low Galactic longitudes (cf. Figs.~\ref{fig:apass_r} and \ref{fig:apass_i}),
which were typically observed during the summer months
when photometric conditions are more prevalent at the telescope.
These remaining fields have nevertheless 
been brought onto the same uniform scale 
by employing an ubercalibration-style scheme,
which minimises the magnitude offsets between stars
located in the overlap regions between neighbouring fields.

An algorithm for achieving this minimisation
has previously been described by~\citet{Glazebrook1994}.
In brief, there are two fundamental quantities to be
minimised between each pair of overlapping exposures, 
denoted by the indices $i$ and $j$. 
Firstly, the mean magnitude difference between stars in the overlap
region $\Delta_{ij}=\langle m_i-m_j\rangle=-\Delta_{ji}$ is a local
constraint. 
Secondly, to ensure the solution does not stray far 
from the existing calibration, 
the difference in zeropoints 
$\Delta\mathrm{ZP}_{ij}=-\Delta\mathrm{ZP}_{ji}$ 
between each pair of exposures must also be minimised.

Minimisation of these two quantities is a linear least squares problem 
because the magnitude $m$ depends linearly on the ZP (Eqn.~\ref{eqn:mag}).
Hence we can find the ZP shift to be applied to each field 
by minimising the sum:
\begin{equation}
   S = \sum_{i=1}^N \sum_{j=1}^N w_{ij} \theta_{ij} (\Delta_{ij} + a_i - a_j)^2,
   \label{eqn:chi2}
\end{equation}
where $i$ denotes an exposure, 
$j$ an overlapping exposure, 
$N$ the number of exposures,
$a_i$ the ZP to solve for 
and $a_j$ the ZP of an overlapping field ($\Delta\mathrm{ZP}_{ij}=a_i-a_j$). 
$w_{ij}$ are weights set to the inverse square of the uncertainty in $\Delta_{ij}$,
and $\theta_{ij}$ is an overlap function 
equal to either 1 if exposures $i$ and $j$ overlap or 0 otherwise. 
Solving for $a_i$ is equivalent to solving $\partial
S/\partial a_i=0$, which gives the matrix equation:
\begin{equation}
   \sum_{j=1}^N A_{ij} a_j = b_j,
   \label{eqn:matrix}
\end{equation}
where 
\begin{eqnarray}
   A_{ij} &=& \delta_{ij} \sum_{k=1}^N w_{jk}\theta_{jk} - w_{ij} \theta_{ij},\label{eqn:aij}\\
   b_i &=& \sum_{j=1}^N w_{ij} \theta_{ij}\Delta_{ji} = -\sum_{j=1}^N w_{ij} \theta_{ij}\Delta_{ij}.\label{eqn:bi}
\end{eqnarray}

We enforce a strong external constraint
on the solution by keeping the zeropoint fixed 
for the 
% 6\,756 
fields which have already been compared
and calibrated against APASS.
We hereafter refer to these fields as \emph{anchors}.
It is asserted that the zeropoints $a_i$ of the anchor fields 
are known and not solved for.  However they do appear in the vector 
$b_j$ as constraints.  In addition to the APASS-based anchors, 
we selected 3\,273 additional anchor fields by hand
to provide additional constraints in regions not covered by APASS.
These extra anchors were deemed to have reliable zeropoints 
based on 
(i) the information contained in the observing logs,
(ii) the stability of the standard star zeropoints during the night, and
(iii) photometricity statistics provided by the Carlsberg Meridian Telescope,
which is located $\sim$500~m from the INT.

We then solved Eqn.~\ref{eqn:matrix} for the $r$ and $i$ bands
separately using the least-squares routine 
in Python's {\sc scipy.sparse} module for sparse matrix algebra.
This provided us with corrected zeropoints for the remaining fields,
which were shifted on average by $+0.02\pm0.11$ in $r$ 
and $+0.01\pm0.12$ in $i$ compared to their provisional calibration.
The outcome is validated in the next section.

We then turned to the global calibration of the \ha\ data.
It is not possible to re-calibrate the narrowband 
in the same way as the broadbands,
because the APASS survey does not offer \ha\ photometry.
We can reasonably assume however,
that the corrections required for $r$ and \ha\ are identical,
much of the time, because the IPHAS data-taking pattern ensured 
that a field's \ha\ and $r$-band exposures
were taken consecutively,
albeit separated by the $\sim$30~s read-out time.
Hence, we have corrected the \ha\ zeropoints 
by re-using the zeropoint adjustments that were derived for the $r$ band
in the earlier steps.
An exception was made for a small subset of fields
% 3\,101 fields
for which our quality-control routines revealed
strong zeropoint variations during the night,
suggesting non-photometric conditions.
In these cases, the \ha\ zeropoints
were adjusted by solving Eqn.~\ref{eqn:matrix}
rather than by simply applying Eqn.~\ref{eqn:zpha}.

@@

Having re-calibrated all fields to the expected APASS accuracy of 3 per cent,
we turned to a different survey, SDSS Data Release 9 \citep{Ahn2012},
to validate the results.
SDSS DR9 includes several strips at low
Galactic latitudes,
providing data across 18 per cent of the fields in our data release.
We cross-matched the IPHAS fields against the subset of
objects marked as reliable stars in the SDSS catalogue\footnote{
We used the CasJobs facility located at http://skyserver.sdss3.org/CasJobs
to obtain photometry from the SDSS {\sc photoprimary} table 
with criteria {\sc type = star}, {\sc clean = 1} and {\sc score $>$ 0.7}.}
cross-matching in much the same way as for APASS,
with the difference of selecting from the  fainter magnitude ranges of 
$15<r_{\rm SDSS}<18.0$ and $14.5<i_{\rm SDSS}<17.5$.
This provided us with a set of 1.2~million cross-matched stars.

>>>>

\begin{figure*}
        \includegraphics[width=1\linewidth]{figures/footprint/footprint_small.png}
        \caption{Survey area showing the footprints
        of all the quality-approved IPHAS fields
        which have been included in this data release.
        The area covered by each field has been coloured black
        with a semi-transparent opacity of 20 per cent,
        such that regions where fields overlap are darker.
        The IPHAS strategy is to observe each field twice
        with a small offset,
        and hence the vast majority of the area 
        is covered twice (dominant grey colour).
        There are small overlaps between all the neighbouring fields
        which can be seen as a honeycomb pattern
        of dark grey lines across the survey area.
        Regions with incomplete data are apparent as white gaps (no data) 
        or in light grey (indicating that one offset is missing).
        The dark vertical strip near $\ell \simeq$125\degr\ 
        is an arbitrary consequence of the tiling pattern,
        which was populated starting from 0h in Right Ascension.}
        \label{fig:footprint}
\end{figure*}

@@

(3) \emph{Seeing.} 
The original survey goal was to obtain data 
at a resolution better than 1.7~arcsec,
as evaluated by measuring the average PSF Full Width at Half Maximum (FWHM)
across the detectors.
This target is currently attained across 86 per cent of the footprint.
To increase the sky area offered by the data release slightly,
we have decided to accept data obtained with FWHM up to 2.5~arcsec.
Fig.~\ref{fig:seeing} presents the distribution
of the PSF FWHM for the approved fields.
In the $r$-band, 90 per cent is better than 1.5~arcsec,
50 per cent is better than 1.1~arcsec,
and 10 per cent is better than 0.8~arcsec.
In \S\ref{sec:catalogue} we will explain
that the photometry compiled in the source catalogue
is normally derived from the field with the
best-available seeing for a given object,
and that the FWHM measurement
is available as a column in the catalogue.

@@

Unlike the broadbands, 
standard-star photometry is not available in the literature 
for the \ha\ passband
and hence there is no formally recognised flux scale for it.
We can specify here, however, 
that the detected flux for Vega in the
IPHAS \ha\ filter is 3.14 magnitudes
less than the flux captured by the much broader
$r$ band \citep{Gonzalez-Solares2008}.
Hence to assure that $(r-\ha)=0$ for Vega,
we set the zeropoint for the narrowband to be:
\begin{equation}
\textsc{magzpt}_{H\alpha} = \textsc{magzpt}_r - 3.14.
\label{eqn:zpha}
\end{equation}
For reference, Table~\ref{tbl:flux} details the flux of Vega
in the IPHAS filter system.
Data on the throughput curves of the bands can be obtained from
the Isaac Newton Group website\footnote{http://catserver.ing.iac.es/filter/list.php?instrument=WFC where the
filters are named WFCH6568, WFCSloanR and WFCSloanI}.

\begin{table}
    \caption{Mean monochromatic flux of Vega in the IPHAS filter system,
             defined as 
             $\langle f_\lambda\rangle~=~\int{f_\lambda(\lambda) S(\lambda) \lambda d \lambda} / \int{S(\lambda)\lambda d \lambda}$,
             where $S(\lambda)$ is the photon response function 
             (which includes atmospheric transmission,
             filter transmission
             and CCD response)
             and $f_\lambda(\lambda)$ is the CALSPEC SED for Vega
             \citep{Bohlin2014}.
             For reference, we also provide the 
             filter equivalent width EW~=~$\int S(\lambda) d\lambda$,
             the mean photon wavelength 
             $\lambda_0~=~\int{S(\lambda)\lambda d\lambda} / \int{S(\lambda)d\lambda}$,
             and the pivot wavelength
             $\lambda_p~=~\sqrt{\int{S(\lambda)\lambda d\lambda} / \int{\frac{S(\lambda)}{\lambda} d\lambda}}$.
             These notations follow the definitions by \citet{Bessell2012}.
             After multiplying $\langle f_\lambda\rangle$ by the EW,
             we find that the detected flux for Vega in \ha\ is 3.14~magnitudes less
             than that received in $r$.
          }
    \label{tbl:flux}
    \begin{center}
    \begin{tabular}{lccrc}
    \toprule
     Filter  & $\langle f_\lambda \rangle$  & \multicolumn{1}{c}{EW} &$\lambda_0$ & $\lambda_p$    \\
             & [erg\,cm$^{-2}$\,s$^{-1}$\,\AA$^{-1}$] & [\AA] & [\AA] & [\AA] \\
    \midrule
$r$ &  $2.47 \times 10^{-9}$  &  785.6 &  6223  &   6211   \\
\ha &  $1.81 \times 10^{-9}$ &   59.6 &  6568  &   6568    \\
$i$  &  $1.30 \times 10^{-9}$  &  759.9  &  7674  &   7661   \\
    \bottomrule
    \end{tabular}
    \end{center}
\end{table}

@@

Despite the best efforts made to obtain a nightly calibration,
large surveys naturally possess field-to-field variations
due to atmospheric changes during the night
and imperfections in the pipeline or the instrument
(e.g. the WFC is known to suffer from sporadic errors
in the timing of exposures).
This is demonstrated in Fig.~\ref{fig:ccd_before},
where we show the combined colour-colour diagram
for the set of 2\,801 quality-approved fields
which are located towards the Galactic anti-centre
($160\degr < \ell < 200\degr$).
The main locus of stars is poorly defined
in the diagram due to the presence of
incorrectly calibrated fields.
The calibration errors can exceed 0.1~mag
and need to be corrected for 
during a global re-calibration procedure.
For comparison, Fig.~\ref{fig:ccd_after}
shows the identical diagram
obtained after re-calibrating.

\begin{figure}
\captionsetup[subfigure]{labelformat=empty}
\begin{subfigure}[b]{\linewidth}
\centering
\includegraphics[width=\textwidth]{figures/caldiagram/ccd-uncalibrated.pdf}
\caption{}
\label{fig:ccd_before}
\end{subfigure}
\begin{subfigure}[b]{\linewidth}
\centering
\includegraphics[width=\textwidth]{figures/caldiagram/ccd-calibrated.pdf}
\caption{}
\label{fig:ccd_after}
\end{subfigure}
\caption{IPHAS $(r-\ha,\ r-i)$ colour-colour diagram
         covering an area of 400~deg$^2$,
         shown before (panel a) and after (panel b)
         re-calibration.
         Both figures were created by combining the stars
         detected across all 2,801 quality-approved fields
         which are located towards the Galactic anti-centre
         ($160\degr<\ell<200\degr$).
         The diagrams are plotted as 2D-histograms
         which show the density of sources
         in bins of 0.01-by-0.01~mag;
         bins containing 1 to 200 sources are coloured red,
         while bins containing more than 400 sources are bright yellow.
         The diagrams include all stars
         brighter than $r<18$ which were
         classified by the pipeline as being `very reliable'
         (to be explained in
         \S\ref{sec:qualitycriteria}).
         The objects seen to fall above the locus of stars
         \emph{after} re-calibration 
         are likely to be genuine \ha\ emission-line objects.}
\end{figure}

Notable past examples of surveys which required
global re-calibration include 
2MASS \citep{Nikolaev2000},
SDSS \citep{Padmanabhan2008}
and the Panoramic Survey Telescope 
And Rapid Response System survey \citep[Pan-STARRS;][]{Schlafly2012},
which all achieved photometry 
that is globally consistent to within 0.01--0.02~mag.

Surveys which observe identical stars at different epochs
can use the repeat measurements to ensure a homogeneous calibration.
For example, 2MASS attained its global calibration
by observing six standard fields each hour, 
allowing zeropoint variations to be tracked 
over very short timescales \citep{Nikolaev2000}.
Alternatively, the SDSS and PanSTARRS surveys could benefit
from revisiting regions in their footprint to 
carry out a so-called \emph{ubercalibration}\footnote{`Ubercalibration'
refers to the name of the code used to re-calibrate SDSS photometry
using repeat observations. 
It is an anglicised version of the German word `\"uberkalibration',
which was reportedly chosen because the initial authors both
had German-sounding names \citep{Finkbeiner2010}.} procedure,
in which multiple measurements of stars at different epochs
are used to fit the calibration parameters
\citep{Ivezic2007,Padmanabhan2008,Schlafly2012}.

@@

We have found the information contained
in these narrow overlap regions to be insufficient
to constrain the calibration parameters.
This is because photometry at the extreme edges of the WFC
-- where neighbouring fields overlap -- 
is prone to systematics at the level of 1--2~per cent.
The cause of these errors is thought to include 
the use of twilight sky flats in the pipeline,
which are known to be imperfect for calibrating stellar photometry 
due to stray light and vignetting \citep[e.g.][]{Manfroid1995}.
Moreover, the illumination correction in the overlap regions
is more affected by a radial geometric distortion in the WFC,
which causes the pixel scale to increase as the edges 
are approached \citep{Gonzalez-Solares2011}.
Although these systematics are reasonably small within a single field,
they can accumulate during the calibration process, causing artificial zeropoint gradients 
across the survey unless controlled by other external constraints.

@@

We have been able to benefit from APASS
(http://www.aavso.org/apass)
to bring most of the survey 
onto a uniform scale.
Since 2009,
APASS has been using two 20~cm-astrographs
to survey the entire sky down to $\sim$17th magnitude
in five filters which include Sloan $r$ and $i$ \citep{Henden2012}.
The most recent catalogue available 
at the time of preparing this work was APASS DR7,
which provides a good coverage across $\sim$half
of the IPHAS footprint.
The overlap regions are shown in Fig.~\ref{fig:apass_r}.
The photometric accuracy of APASS is currently estimated 
to be at the level of 3 per cent,
which is significantly better 
than the original nightly calibrations of IPHAS
which are only accurate to $\sim$10 per cent
when compared to APASS (Table~\ref{tbl:offsets_before}).
APASS achieves its uniform accuracy 
by measuring each star at least two times in photometric conditions,
along with ample standard fields,
benefiting from the large 3-by-3 degrees field of view of its detectors.

@@

The resulting set of 220,000 cross-matched stars were then used 
to derive APASS-to-IPHAS magnitude transformations
using a linear least-squares fitting routine, 
which iteratively removed $3\sigma$-outliers to improve the fit.
The solution converged to:
\begin{align} 
r_{\rm IPHAS} = r_{\rm APASS} - 0.121 + 0.032(r-i)_{\rm APASS} \label{eqn:apass_r} \\
i_{\rm IPHAS} = i_{\rm APASS} - 0.364 + 0.006(r-i)_{\rm APASS} \label{eqn:apass_i}
\end{align}
The root mean square (rms) residuals of these transformations 
are 0.041 and 0.051, respectively.
The small colour terms in the equations
indicate that the IPHAS and APASS broadband filters 
are very similar.
The transformations include a large fixed offset,
but this is simply due to the fact that 
APASS magnitudes are given in the AB system
and IPHAS uses Vega-based magnitudes.
Separate transformations were derived for sightlines 
with varying extinction properties to investigate the robustness
of the transformations with respect to different reddening regimes.
This sensitivity was found to be insignificant, in keeping with
the scarcity of heavily-reddened objects at $r<16$.

Having transformed APASS magnitudes into the IPHAS system,
we then computed the median magnitude offset 
for each field which contained at least 30 cross-matched stars.
This was achieved for 48 per cent of our fields (shown in Figs.~\ref{fig:apass_r} and \ref{fig:apass_i}).
The offsets follow a near-Gaussian distribution
with mean and sigma $0.014\pm0.104$~mag in $r$
and $0.007\pm0.108$~mag in $i$ (Table~\ref{tbl:offsets_before}).
A total of 4,596 IPHAS fields showed a median offset
exceeding $\pm$0.03~mag in either $r$ or $i$
when compared to APASS.

@@

We then applied the most important step
in our re-calibration scheme,
which is to adjust the provisional zeropoints
of these 4,596 aberrant fields
such that their offset is brought to zero.
This allowed the mean IPHAS-to-APASS offset 
to be brought down to $0.000\pm0.011$~mag in both $r$ and $i$
(Table~\ref{tbl:offsets_after}).
The procedure of fitting magnitude transformations and
correcting the IPHAS zeropoints was repeated a few times to ensure 
convergence, which was closely approached after the first iteration.

@@

At the time of writing, the APASS catalogue did not provide 
sufficient coverage for 7,359 of our fields.
Fortunately, these fields are located mainly 
at low Galactic longitudes (cf. Figs.~\ref{fig:apass_r} and \ref{fig:apass_i}),
which were typically observed during the summer months
when photometric conditions are more prevalent at the telescope.
These remaining fields have nevertheless 
been brought onto the same uniform scale 
by employing an ubercalibration-style scheme,
which minimises the magnitude offsets between stars
located in the overlap regions between neighbouring fields.

An algorithm for achieving this minimisation
has previously been described by~\citet{Glazebrook1994}.
In brief, there are two fundamental quantities to be
minimised between each pair of overlapping exposures, 
denoted by the indices $i$ and $j$. 
First, the mean magnitude difference between stars in the overlap
region $\Delta_{ij}=\langle m_i-m_j\rangle=-\Delta_{ji}$ is a local
constraint. 
Second, to ensure the solution does not stray far 
from the existing calibration, 
the difference in zeropoints 
$\Delta\mathrm{ZP}_{ij}=-\Delta\mathrm{ZP}_{ji}$ 
between each pair of exposures must also be minimised.

Minimisation of these two quantities is a linear least squares problem 
because the magnitude $m$ depends linearly on the ZP (Eqn.~\ref{eqn:mag}).
Hence we can find the ZP shift to be applied to each field 
by minimising the sum:
\begin{equation}
   S = \sum_{i=1}^N \sum_{j=1}^N w_{ij} \theta_{ij} (\Delta_{ij} + a_i - a_j)^2,
   \label{eqn:chi2}
\end{equation}
where $i$ denotes an exposure, 
$j$ an overlapping exposure, 
$N$ the number of exposures,
$a_i$ the ZP to solve for,
and $a_j$ the ZP of an overlapping field ($\Delta\mathrm{ZP}_{ij}=a_i-a_j$). 
$w_{ij}$ are weights set to the inverse square of the uncertainty in $\Delta_{ij}$,
and $\theta_{ij}$ is an overlap function 
equal to either 1 if exposures $i$ and $j$ overlap or 0 otherwise. 
Solving for $a_i$ is equivalent to solving $\partial
S/\partial a_i=0$, which gives the matrix equation:
\begin{equation}
   \sum_{j=1}^N A_{ij} a_j = b_j,
   \label{eqn:matrix}
\end{equation}
where 
\begin{eqnarray}
   A_{ij} &=& \delta_{ij} \sum_{k=1}^N w_{jk}\theta_{jk} - w_{ij} \theta_{ij},\label{eqn:aij}\\
   b_i &=& \sum_{j=1}^N w_{ij} \theta_{ij}\Delta_{ji} = -\sum_{j=1}^N w_{ij} \theta_{ij}\Delta_{ij}.\label{eqn:bi}
\end{eqnarray}

We enforce a strong external constraint
on the solution by keeping the zeropoint fixed 
for the 
% 6\,756 
fields which have already been compared
and calibrated against APASS.
We hereafter refer to these fields as \emph{anchors}.
It is asserted that the zeropoints $a_i$ of the anchor fields 
are known and not solved for.  However they do appear in the vector 
$b_j$ as constraints.  In addition to the APASS-based anchors, 
we selected 3,273 additional anchor fields by hand
to provide additional constraints in regions not covered by APASS.
These extra anchors were deemed to have reliable zeropoints 
based on 
(i) the information contained in the observing logs,
(ii) the stability of the standard star zeropoints during the night, and
(iii) photometricity statistics provided by the Carlsberg Meridian Telescope,
which is located $\sim$500~m from the INT.

We then solved Eqn.~\ref{eqn:matrix} for the $r$ and $i$ bands
separately using the least-squares routine 
in Python's {\sc scipy.sparse} module for sparse matrix algebra.
This provided us with corrected zeropoints for the remaining fields,
which were shifted on average by $+0.02\pm0.11$ in $r$ 
and $+0.01\pm0.12$ in $i$ compared to their provisional calibration.

We then turned to the global calibration of the \ha\ data.
It is not possible to re-calibrate the narrowband 
in the same way as the broadbands,
because the APASS survey does not offer \ha\ photometry.
We can reasonably assume, however,
that the corrections required for $r$ and \ha\ are identical,
much of the time, because the IPHAS data-taking pattern ensured 
that a field's \ha\ and $r$-band exposures
were taken consecutively,
albeit separated by a $\sim$30~s read-out time.
Hence, we have corrected the \ha\ zeropoints 
by re-using the zeropoint adjustments that were derived for the $r$ band
in the earlier steps.
An exception was made for 3,101 fields
for which our quality-control routines revealed
strong zeropoint variations during the night,
suggesting non-photometric conditions.
In these cases, the \ha\ zeropoints
were adjusted by solving Eqn.~\ref{eqn:matrix}
rather than by simply applying Eqn.~\ref{eqn:zpha}.

@@

Having re-calibrated all fields to the expected APASS accuracy of 3 per cent,
we turned to a different survey, SDSS Data Release 9 \citep{Ahn2012},
to validate the results.
SDSS DR9 includes several strips at low
Galactic latitudes,
providing data across 18 per cent of the fields in our data release.
We cross-matched the IPHAS fields against the subset of
objects marked as reliable stars in the SDSS catalogue\footnote{
We used the CasJobs facility located at http://skyserver.sdss3.org/CasJobs
to obtain photometry from the SDSS {\sc photoprimary} table 
with criteria {\sc type = star}, {\sc clean = 1} and {\sc score $>$ 0.7}.}
in much the same way as for APASS,
with the difference of selecting from the  fainter magnitude ranges of 
$15<r_{\rm SDSS}<18.0$ and $14.5<i_{\rm SDSS}<17.5$.
This provided us with a set of 1.2~million cross-matched stars.
